---
layout: post
title: Heteroscedastic Gaussian process regression
date: 2020-06-22
math: true
keywords:
  - Bayesian statistics
  - Gaussian process
  - R
  - Stan
  - data science
  - heteroscedasticity
  - regression
---

```{r, echo = FALSE, message = FALSE}
library(rstan)
library(tidybayes)
library(tidyverse)

source('../_scripts/2020-06-22-gaussian-process/common.R')

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
theme_set(theme_minimal(base_size = 14))

set.seed(42)
```

Gaussian process regression is a nonparametric Bayesian technique for modeling
relationships between variables of interest. The vast flexibility and rigor
mathematical foundation of this approach make it the default choice in many
problems involving small to medium-sized data sets.

In what follows, we illustrate how Gaussian process regression can be used in
practice. To make the case even more compelling, we shall consider a setting
where linear regression would be inadequate. The focus will be not on getting
the job done as fast as possible but on the learning process.

Consider the following example taken from [_Semiparametric Regression_][SR] by
Ruppert _et al._:

```{r data, include = FALSE}
data(lidar, package = 'SemiPar')
data <- lidar %>%
  transmute(x = (range - min(range)) / diff(range(range)),
            y = logratio)
lidar %>%
  ggplot(aes(range, logratio)) +
  geom_point(size = 1) +
  labs(x = 'Distance',
       y = 'Response')
```

![](/assets/images/2020-06-22-gaussian-process/data-1.svg)

The figure shows 221 observations collected in a [light detection and
ranging][LIDAR] experiment. Each observation can be interpreted as the sum of
the true underlying response at the corresponding distance and random noise. It
can be clearly seen that the variance of the noise varies with the distance: the
spread is substantially larger toward the left-hand side. This phenomenon is
known as heteroscedasticity. Homoscedasticity (the absence of
heteroscedasticity) is one of the key assumptions of linear regression. Applying
linear regression to the above problem would yield suboptimal results. The
estimates of the regression coefficients would still be unbiased; however, the
standard errors of the coefficients would be incorrect and hence misleading. A
different modeling technique is needed in this case.

The above data set will be our running example. For formally and slightly more
generally, we assume that there is a data set of $$m$$ observations:

$$
\left\{
  (\mathbf{x}_i, y_i): \,
  \mathbf{x}_i \in \mathbb{R}^d; \,
  y_i \in \mathbb{R}; \,
  i = 1, \dots, m
\right\}
$$

where the independent variable, $$\mathbf{x}$$, is $$d$$-dimensional, and the
dependent variable, $$y$$, is scalar. In the running example, $$d$$ is 1, and
$$m$$ is 221. It is time for modeling.

# Model

To begin with, consider the following model with additive noise:

$$
y_i = f(\mathbf{x}_i) + \epsilon_i, \text{ for } i = 1, \dots, m.
$$

In the above, $$f: \mathbb{R}^d \to \mathbb{R}$$ represents the true but unknown
underlying function, and $$\epsilon_i$$ represents the perturbation of the
$$i$$th observation by random noise. In the classical linear-regression setting,
the unknown function is modeled as a linear combination of (arbitrary
transformations of) the $$d$$ covariates. Instead of assuming any particular
functional form, we put a Gaussian-process prior on the function:

$$
f(\mathbf{x}) \sim \text{Gaussian process}\left( 0, k(\mathbf{x}, \mathbf{x}') \right).
$$

The above notation means that, before observing any data, the function is a draw
from a Gaussian process with zero mean and a covariance function $$k$$. The
covariance function dectates the degree of correlation between two arbitrary
locations $$\mathbf{x}$$ and $$\mathbf{x}'$$ in $$\mathbb{R}^d$$. For instance,
a frequent choice for $$k$$ is the squared-exponential covariance function:

$$
k(\mathbf{x}, \mathbf{x}')
= \sigma_\text{process}^2 \exp\left( -\frac{\|\mathbf{x} - \mathbf{x}'\|_2^2}{2 \, \ell_\text{process}^2} \right)
$$

where $$\|\cdot\|_2$$ stands for the Euclidean norm, $$\sigma_\text{process}^2$$
is the variance (to see this, substitute $$\mathbf{x}$$ for $$\mathbf{x}'$$),
and $$\ell_\text{process}$$ is known as the length scale. While the variance
parameter is intuitive, the length-scale one requires an illustration. The
parameter controls the speed with which the correlation fades with the distance.
The following figure shows 10 random draws for $$\ell_\text{process} = 0.1$$:

```{r prior-short, include = FALSE}
prior_plot(ell_process = 0.1)
```

![](/assets/images/2020-06-22-gaussian-process/prior-short-1.svg)

With $$\ell_\text{process} = 0.5$$, the behavior change to the following:

```{r prior-long, include = FALSE}
prior_plot(ell_process = 0.5)
```

![](/assets/images/2020-06-22-gaussian-process/prior-long-1.svg)

It can be seen that, in the second case with a large length scale, it takes a
greater distance for the function to change.

The error terms are modeled as independent identically distributed Gaussian
random variables:

$$
\epsilon_i \sim \text{Gaussian}\left( 0, \sigma_\text{noise}^2 \right),
\text{ for } i = 1, \dots, m.
$$

Now,

$$
\begin{align}
& f(\mathbf{x}) \sim \text{Gaussian process}\left(\mathbf{0}, k(\mathbf{x}, \mathbf{x}')\right); \\
& \epsilon_i \sim \text{Gaussian}\left(0, \sigma^2_\text{noise}\right),
\text{ for } i = 1, \dots, m.
\end{align}
$$

# Acknowledgments

I would like to thank Mattias Villani.

# References

* Carl Rasmussen _et al._, [_Gaussian Processes for Machine
  Learning_][GPML], the MIT Press, 2006.

* David Ruppert _et al._, [_Semiparametric Regression_][SR], Cambridge
  University Press, 2003.

* Mattias Villani, “[Advanced Bayesian learning][ABL],” Stockholm
  University, 2020.

[ABL]: https://github.com/mattiasvillani/AdvBayesLearnCourse
[GPML]: http://www.gaussianprocess.org/gpml
[LIDAR]: https://en.wikipedia.org/wiki/Lidar
[SR]: http://www.stat.tamu.edu/~carroll/semiregbook
