---
layout: post
title: Heteroscedastic regression by virtue of Gaussian processes
date: 2020-06-22
math: true
keywords:
  - Bayesian statistics
  - Gaussian process
  - R
  - Stan
  - data science
  - heteroscedasticity
  - regression
---

```{r, echo = FALSE, message = FALSE}
library(rstan)
library(tidybayes)
library(tidyverse)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
theme_set(theme_minimal(base_size = 14))

set.seed(42)
```

Consider the following example taken from [_Semiparametric Regression_][SR] by
Ruppert _et al._:

```{r data, include = FALSE}
data(lidar, package = 'SemiPar')
data <- lidar %>%
  transmute(x = (range - min(range)) / diff(range(range)),
            y = logratio)
lidar %>%
  ggplot(aes(range, logratio)) +
  geom_point(size = 1) +
  labs(x = 'Distance',
       y = 'Response')
```

![](/assets/images/2020-06-22-gaussian-process/data-1.svg)

The figure shows 221 observations collected in a [light detection and
ranging][LIDAR] experiment. Each observation can be interpreted as the sum of
the true underlying response at the corresponding distance and random noise. It
can be clearly seen that the variance of the noise varies with the distance: the
spread is substantially larger toward the left-hand side. This phenomenon is
known as heteroscedasticity. Homoscedasticity (the absence of
heteroscedasticity) is one of the key assumptions of linear regression. Applying
linear regression to the above problem would yield suboptimal results.
More specifically, the estimates of the regression coefficients would still be
unbiased; however, the standard errors of the coefficients would be incorrect
and hence misleading.

# Acknowledgments

I would like to thank Mattias Villani.

# References

* Carl Rasmussen _et al._, [_Gaussian Processes for Machine
  Learning_][GPML], the MIT Press, 2006.

* David Ruppert _et al._, [_Semiparametric Regression_][SR], Cambridge
  University Press, 2003.

* Mattias Villani, “[Advanced Bayesian learning][ABL],” Stockholm
  University, 2020.

[ABL]: https://github.com/mattiasvillani/AdvBayesLearnCourse
[GPML]: http://www.gaussianprocess.org/gpml
[LIDAR]: https://en.wikipedia.org/wiki/Lidar
[SR]: http://www.stat.tamu.edu/~carroll/semiregbook
